Lossy compression of DNA sequencing quality data

### 2018-03-03 peerj_manuscript/manuscript.tex
removed some blank space
changed many citations from \cite to \citep, which is more appropriate for PeerJ and how we actually cite references

Advantageous use of quality values has been long
recognize~\cite{Bonfield:1995kq}, and quality values can be used
throughout bioinformatics pipelines.  Quality values comprise a
standard component of \textsc{fastq} files~\cite{Cock:2010ve}, a very
common format for sequence read data. At the level of the sequence
read, the probability of error for each base-call is typically
represented by a \textsc{phred} quality value, which is defined as $Q
= -10\,log_{10}P$~\cite{Ewing:1998ly}. Depending on the sequencing
technology, these quality values can range from 0 to 93, and are
represented with the \textsc{ascii} characters 33 to 126 (with some
offset). There is a single quality value per base-call for Illumina
sequence reads.

Among the most fundamental uses of sequence quality values is as part
of the quality assessment and quality control (\textsc{qa/qc})
processes prior to subsequent analysis steps. Quality control based on
quality values generally includes two operations:
\textit{i}.~filtering, which removes reads that on the whole do not
meet arbitrary quality standards, thus reducing the total number of
reads; and \textit{ii}.~trimming of low-quality base-calls from reads,
which reduces the length of reads trimmed. Quality values can be used
by genome assembly software to produce better
assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
alignment programs, such as Bowtie2~\cite{Langmead:2012rw}, use
quality values to weight mismatches between read and reference
sequences. Software for detecting single nucleotide polymorphisms
(\textsc{snp}s) can use quality values~\cite[e.g.,][]{McKenna:2010bh},
and identified \textsc{snp}s with high-quality calls are deemed more
reliable than those with low-quality calls, particularly in
low-coverage regions.

@@

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
bioinformatic analyses. From this perspective, some loss of
information is deemed acceptable if the inferences from analyses are
relatively unaffected. Sequence reads generated by instruments such as
an Illumina HiSeq, the focus of this research due to its preponderance
in genomics, are characterized by having relatively few insertion and
deletion errors, but substitution (miscall) errors are much more
frequent and have context-specific patterns. These errors are
non-uniformly distributed over the read length (e.g., error rates
increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and 32.8 --
67.9\% of reads have low-quality tails at the 3$^{\prime}$
end~\cite{Minoche:2011km}). Recognizing these properties of Illumina
sequence reads motivates our exploration of three general classes of
lossy compression methods -- binning, modeling, and profiling -- and
our consideration of an exemplar of each class. Some studies have
evaluated the effects of lossy compression on identifying variants
within a data set~\cite{Canovas:2014fr,janin2013adaptive}. Here we
describe our research investigating lossy compression of sequence read
quality values, specifically those associated with Illumina
instruments, with the objective to provide some perspective on several
strategies rather than to develop robust high-quality software for
use. We assess the effects of quality value information loss resulting
from compression on \textsc{dna} sequence data analyses, including
read preprocessing (filtering and trimming), genome assembly, and read
mapping.
\section*{Methods}

@@

Quality values can be binned, and the minimum number of bins that
allows for any distinction among quality values is two; i.e., two
categories of quality: ``good'' and ``bad''. We implement 2-bin
encoding by setting a quality value threshold empirically determined
by the distribution of quality values across reads. Base-calls are
marked ``bad'' if their quality value falls below the first quartile
minus 1.5 $\times$ the interquartile range (\textsc{iqr}), which is
the difference between the first and third quartile; 1.5 $\times$
\textsc{iqr} is the value used by Tukey's box
plot~\cite{mcgill1978variations}. The main benefit of this approach is
that it is completely data-dependent, and no assumptions regarding the
distribution of the quality values need to be made. We adopt the
quality value assignments of the read preprocessing tool
Sickle~\cite{sickle}, which correspond to the 40 for ``good'' and 10
for ``bad''.

With 2-bin encoding, binary encoding is possible, allowing us to use a
single bit to represent the quality of a base instead of the standard
8 bits used to store quality values in \textsc{ascii}. An additional
possible benefit of 2-bin encoding is for increased adjacency of
identical values and repeating patterns, properties that may increase
effectiveness of subsequent compression using established
algorithms~\cite{HUFFMAN:1952nr,Ziv77auniversal,
  DBLP:journals/tit/ZivL78}, although this potential benefit is not
evaluated in this study.

@@

Previous work by others \cite{Wan:2012kq} describe three similar lossy
compression strategies based on binning the base error probability
distribution: UniBinning, Truncating, and LogBinning. UniBinning
evenly splits the error probability distribution into a user-defined
number of partitions. Truncating treats a user-defined number of
highest quality values as a single bin. LogBinning works similarly to
UniBinning, except it uses the \emph{log} of the error probability
distribution, which effectively bins the \textsc{ascii} quality values
evenly. The 2-bin encoding examined here can be viewed as something
like a combination of LogBinning and Truncating in that we place the
highest quality values (as defined above) from the distribution of log
error probability values into a single bin.

@@

Sets of strings representing quality values show similar trends over
their length, and it is possible to identify common patterns in the
data and use them as reference profiles to approximate individual
sequences of quality values. Here we use $k$-means clustering, a
vector quantization method that partitions a set of samples into $k$
sets that minimize within-cluster sum of
squares~\cite{macqueen1967some}. We sampled 1 $\times$ 10$^{4}$ reads
at random and computed cluster centers as read quality profiles using
a heuristic iterative refinement approach that quickly converges to a
locally optimal minimum~\cite{hartigan1979algorithm}. All reads are
evaluated and the nearest quality profile in Euclidean space being
assigned to every read as their compressed representation.  The
compressed quality file therefore consists of an index enumerating the
$k$ quality profiles, and a binary part containing the assigned
quality profile index for each read.  Although this approach is not
able to capture some read-specific differences in quality values, it
does approximate the overall patterns in quality values. An example of
128 quality profiles is shown in Figure \ref{fig:profiles_128}.

@@

We used several Illumina sequence read data sets taken from data from
the \textsc{gage} (Genome Assembly Gold-Standard
Evaluations)~\cite{Salzberg:2012rc}, except as noted. These data sets
are as follows.

@@

For comparison to other developed methods for lossy compression of
quality values we analyzed the reference data with
\textsc{QualComp}~\cite{Ochoa:2013rt}, and Quality Values Zip
(\textsc{qvz)}~\cite{Malysa01102015}, and Read Quality Sparsifier
(\textsc{rqs})~\cite{DBLP:conf/recomb/YuYB14}

@@

We secondarily compress all otherwise compressed data sets using the
Burrows-Wheeler algorithm~\cite{bwt} via \textsc{bz}ip2, and it is the
results of these subsequent compressions that we report. Although
other algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,
  DBLP:journals/tit/ZivL78} might be particularly effective for
subsequent compression for some compressed data representations
examined, we not explored such possibilities here, and instead
restrict ourselves to \textsc{bz}ip2 because it is commonly used and
readily available. Similarly, it may very well be possible to obtain
increased secondary compression by reordering (e.g., sorting)
compressed representations (e.g., sets of bin designations, models, or
profiles), but these operations have also not been explored here.

@@

As a measure of compression effectiveness, which reflects the sum of
fixed and marginal costs, we use bits/base-call, defined as the size
of the compressed representation of quality values (in bits) divided
by the number of quality values represented. As measures of
information loss we use three loss functions following previous
work~\cite{Malysa01102015}: mean squared error (\textsc{mse}), defined
as $\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$; mean absolute deviation
(\textsc{mad}), defined as $\frac{1}{n}\sum_{i=1}^{n}{|Q_i'-Q_i|}$;
and mean of the values derived from part of the formula defining the
Cauchy--Lorentz distribution (mean Cauchy--Lorentz loss), defined as
$\frac{1}{n}\sum_{i=1}^{n}{log_{2}(1 + |Q_i'-Q_i|)}$, where $n$ is the
number of quality values, $Q_i'$ is the compressed/decompressed
quality value, and $Q_i$ is the original quality value associated with
position $i$.

We evaluate effects of information loss from quality value compression
on quality control steps of trimming and read filtering, which were
performed using Sickle~\cite{sickle}, and make comparison to original
uncompressed data. Sickle starts at the ends of the read and uses a
sliding window (0.1 $\times$ the read length) to find locations where
the mean quality in the window falls below a given threshold (20 by
default). The sequence is then trimmed from this position to the read
end. If the trimmed sequence length is less than a certain threshold
(20 by default), then the sequence is discarded.

We evaluate effects of information loss from quality value compression
on \emph{de novo} genome assembly performance using contiguity
statistics, log average read probability
(\textsc{lap})~\cite{Ghodsi:2013hb}, and a collection of
reference-based metrics. The contiguity statistics include N50, which
is defined as the median contig size (the length of largest contig $c$
such that the total size of the contigs larger than $c$ exceeds half
of the assembly size) and corrected N50, which is the recalculated N50
size after the contigs are broken apart at locations of errors. The
\textsc{lap} score can be viewed as a log-likelihood score, where a
larger value is better. We use a script provided by \textsc{gage}
reference-based evaluation to count single nucleotide polymorphisms
(\textsc{snp}s), relocations, translations, and inversions. The
reference-based metrics are normalized by the length of the assembly
to facilitate comparison. For the genome assembly we used software
that makes use of quality values in the assembly process:
\textsc{allpaths-lg}~\cite{Gnerre:2011kx} version r50191 with default
settings and 32 threads.

@@

As an additional point of reference regarding compression
effectiveness, the recently published
Read-Quality-Sparsifier~\cite{DBLP:conf/recomb/YuYB14} achieved
best-case compression of 0.254 bits/base-call on \textit{Homo sapiens}
chromosome 21 data (read lengths ranging from 50--110 nt), and a mean
compression of 1.841 bits/base-call; both values are larger than all
of the profile results and for some results of other methods presented
here (Figure~\ref{fig:mse_vs_bpbp}).

>>>>

Advantageous use of quality values has been long
recognize~\citep{Bonfield:1995kq}, and quality values can be used
throughout bioinformatics pipelines.  Quality values comprise a
standard component of \textsc{fastq} files~\citep{Cock:2010ve}, a very
common format for sequence read data. At the level of the sequence
read, the probability of error for each base-call is typically
represented by a \textsc{phred} quality value, which is defined as $Q
= -10\,log_{10}P$~\citep{Ewing:1998ly}. Depending on the sequencing
technology, these quality values can range from 0 to 93, and are
represented with the \textsc{ascii} characters 33 to 126 (with some
offset). There is a single quality value per base-call for Illumina
sequence reads.

Among the most fundamental uses of sequence quality values is as part
of the quality assessment and quality control (\textsc{qa/qc})
processes prior to subsequent analysis steps. Quality control based on
quality values generally includes two operations:
\textit{i}.~filtering, which removes reads that on the whole do not
meet arbitrary quality standards, thus reducing the total number of
reads; and \textit{ii}.~trimming of low-quality base-calls from reads,
which reduces the length of reads trimmed. Quality values can be used
by genome assembly software to produce better
assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
alignment programs, such as Bowtie2~\citep{Langmead:2012rw}, use
quality values to weight mismatches between read and reference
sequences. Software for detecting single nucleotide polymorphisms
(\textsc{snp}s) can use quality values~\cite[e.g.,][]{McKenna:2010bh},
and identified \textsc{snp}s with high-quality calls are deemed more
reliable than those with low-quality calls, particularly in
low-coverage regions.

@@

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
bioinformatic analyses. From this perspective, some loss of
information is deemed acceptable if the inferences from analyses are
relatively unaffected. Sequence reads generated by instruments such as
an Illumina HiSeq, the focus of this research due to its preponderance
in genomics, are characterized by having relatively few insertion and
deletion errors, but substitution (miscall) errors are much more
frequent and have context-specific patterns. These errors are
non-uniformly distributed over the read length (e.g., error rates
increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and 32.8 --
67.9\% of reads have low-quality tails at the 3$^{\prime}$
end~\citep{Minoche:2011km}). Recognizing these properties of Illumina
sequence reads motivates our exploration of three general classes of
lossy compression methods -- binning, modeling, and profiling -- and
our consideration of an exemplar of each class. Some studies have
evaluated the effects of lossy compression on identifying variants
within a data set~\citep{Canovas:2014fr,janin2013adaptive}. Here we
describe our research investigating lossy compression of sequence read
quality values, specifically those associated with Illumina
instruments, with the objective to provide some perspective on several
strategies rather than to develop robust high-quality software for
use. We assess the effects of quality value information loss resulting
from compression on \textsc{dna} sequence data analyses, including
read preprocessing (filtering and trimming), genome assembly, and read
mapping.
\section*{Methods}

@@

Quality values can be binned, and the minimum number of bins that
allows for any distinction among quality values is two; i.e., two
categories of quality: ``good'' and ``bad''. We implement 2-bin
encoding by setting a quality value threshold empirically determined
by the distribution of quality values across reads. Base-calls are
marked ``bad'' if their quality value falls below the first quartile
minus 1.5 $\times$ the interquartile range (\textsc{iqr}), which is
the difference between the first and third quartile; 1.5 $\times$
\textsc{iqr} is the value used by Tukey's box
plot~\citep{mcgill1978variations}. The main benefit of this approach is
that it is completely data-dependent, and no assumptions regarding the
distribution of the quality values need to be made. We adopt the
quality value assignments of the read preprocessing tool
Sickle~\citep{sickle}, which correspond to the 40 for ``good'' and 10
for ``bad''.

With 2-bin encoding, binary encoding is possible, allowing us to use a
single bit to represent the quality of a base instead of the standard
8 bits used to store quality values in \textsc{ascii}. An additional
possible benefit of 2-bin encoding is for increased adjacency of
identical values and repeating patterns, properties that may increase
effectiveness of subsequent compression using established
algorithms~\citep{HUFFMAN:1952nr,Ziv77auniversal,
  DBLP:journals/tit/ZivL78}, although this potential benefit is not
evaluated in this study.

@@

Previous work by others \citep{Wan:2012kq} describe three similar lossy
compression strategies based on binning the base error probability
distribution: UniBinning, Truncating, and LogBinning. UniBinning
evenly splits the error probability distribution into a user-defined
number of partitions. Truncating treats a user-defined number of
highest quality values as a single bin. LogBinning works similarly to
UniBinning, except it uses the \emph{log} of the error probability
distribution, which effectively bins the \textsc{ascii} quality values
evenly. The 2-bin encoding examined here can be viewed as something
like a combination of LogBinning and Truncating in that we place the
highest quality values (as defined above) from the distribution of log
error probability values into a single bin.

@@

Sets of strings representing quality values show similar trends over
their length, and it is possible to identify common patterns in the
data and use them as reference profiles to approximate individual
sequences of quality values. Here we use $k$-means clustering, a
vector quantization method that partitions a set of samples into $k$
sets that minimize within-cluster sum of
squares~\citep{macqueen1967some}. We sampled 1 $\times$ 10$^{4}$ reads
at random and computed cluster centers as read quality profiles using
a heuristic iterative refinement approach that quickly converges to a
locally optimal minimum~\citep{hartigan1979algorithm}. All reads are
evaluated and the nearest quality profile in Euclidean space being
assigned to every read as their compressed representation.  The
compressed quality file therefore consists of an index enumerating the
$k$ quality profiles, and a binary part containing the assigned
quality profile index for each read.  Although this approach is not
able to capture some read-specific differences in quality values, it
does approximate the overall patterns in quality values. An example of
128 quality profiles is shown in Figure \ref{fig:profiles_128}.

@@

We used several Illumina sequence read data sets taken from data from
the \textsc{gage} (Genome Assembly Gold-Standard
Evaluations)~\citep{Salzberg:2012rc}, except as noted. These data sets
are as follows.

@@

For comparison to other developed methods for lossy compression of
quality values we analyzed the reference data with
\textsc{QualComp}~\citep{Ochoa:2013rt}, and Quality Values Zip
(\textsc{qvz)}~\citep{Malysa01102015}, and Read Quality Sparsifier
(\textsc{rqs})~\citep{DBLP:conf/recomb/YuYB14}

@@

We secondarily compress all otherwise compressed data sets using the
Burrows-Wheeler algorithm~\citep{bwt} via \textsc{bz}ip2, and it is the
results of these subsequent compressions that we report. Although
other algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,
  DBLP:journals/tit/ZivL78} might be particularly effective for
subsequent compression for some compressed data representations
examined, we not explored such possibilities here, and instead
restrict ourselves to \textsc{bz}ip2 because it is commonly used and
readily available. Similarly, it may very well be possible to obtain
increased secondary compression by reordering (e.g., sorting)
compressed representations (e.g., sets of bin designations, models, or
profiles), but these operations have also not been explored here.

@@

As a measure of compression effectiveness, which reflects the sum of
fixed and marginal costs, we use bits/base-call, defined as the size
of the compressed representation of quality values (in bits) divided
by the number of quality values represented. As measures of
information loss we use three loss functions following previous
work~\citep{Malysa01102015}: mean squared error (\textsc{mse}), defined
as $\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$; mean absolute deviation
(\textsc{mad}), defined as $\frac{1}{n}\sum_{i=1}^{n}{|Q_i'-Q_i|}$;
and mean of the values derived from part of the formula defining the
Cauchy--Lorentz distribution (mean Cauchy--Lorentz loss), defined as
$\frac{1}{n}\sum_{i=1}^{n}{log_{2}(1 + |Q_i'-Q_i|)}$, where $n$ is the
number of quality values, $Q_i'$ is the compressed/decompressed
quality value, and $Q_i$ is the original quality value associated with
position $i$.

We evaluate effects of information loss from quality value compression
on quality control steps of trimming and read filtering, which were
performed using Sickle~\citep{sickle}, and make comparison to original
uncompressed data. Sickle starts at the ends of the read and uses a
sliding window (0.1 $\times$ the read length) to find locations where
the mean quality in the window falls below a given threshold (20 by
default). The sequence is then trimmed from this position to the read
end. If the trimmed sequence length is less than a certain threshold
(20 by default), then the sequence is discarded.

We evaluate effects of information loss from quality value compression
on \emph{de novo} genome assembly performance using contiguity
statistics, log average read probability
(\textsc{lap})~\citep{Ghodsi:2013hb}, and a collection of
reference-based metrics. The contiguity statistics include N50, which
is defined as the median contig size (the length of largest contig $c$
such that the total size of the contigs larger than $c$ exceeds half
of the assembly size) and corrected N50, which is the recalculated N50
size after the contigs are broken apart at locations of errors. The
\textsc{lap} score can be viewed as a log-likelihood score, where a
larger value is better. We use a script provided by \textsc{gage}
reference-based evaluation to count single nucleotide polymorphisms
(\textsc{snp}s), relocations, translations, and inversions. The
reference-based metrics are normalized by the length of the assembly
to facilitate comparison. For the genome assembly we used software
that makes use of quality values in the assembly process:
\textsc{allpaths-lg}~\citep{Gnerre:2011kx} version r50191 with default
settings and 32 threads.

@@

As an additional point of reference regarding compression
effectiveness, the recently published
Read-Quality-Sparsifier~\citep{DBLP:conf/recomb/YuYB14} achieved
best-case compression of 0.254 bits/base-call on \textit{Homo sapiens}
chromosome 21 data (read lengths ranging from 50--110 nt), and a mean
compression of 1.841 bits/base-call; both values are larger than all
of the profile results and for some results of other methods presented
here (Figure~\ref{fig:mse_vs_bpbp}).
